---
title: "Group B — Menstrual Cycle Analysis"
subtitle: "Cytokines, Immune Cells, and Birth Control"
format:
  pdf:
    toc: true
    number-sections: true
documentclass: article
geometry: margin=1in
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)     # for tidy model outputs
library(knitr)     # for kable tables
data_dir <- here::here("data")
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This document explores the **Group B (menstrual cycle)** dataset. We'll:

1.  Load and join the data
2.  Make summary tables with p-values
3.  Visualise cytokine and immune cell differences by birth control and time point
4.  Run regression models

---

# 1. Load and Join Data

```{r load-data}
# --- Sample IDs ---
sample_ids <- read_csv(file.path(data_dir, "group_b_menstruation",
                                  "00_sample_ids_period.csv"),
                       show_col_types = FALSE)

# --- Participant metadata ---
metadata <- read_csv(file.path(data_dir, "group_b_menstruation",
                                "01_participant_metadata_period.csv"),
                     show_col_types = FALSE)

# --- Luminex cytokines (long format) ---
luminex <- read_csv(file.path(data_dir, "group_b_menstruation",
                               "02_luminex_period.csv"),
                    show_col_types = FALSE)

# --- Flow cytometry ---
flow <- read_csv(file.path(data_dir, "group_b_menstruation",
                            "03_flow_cytometry_period.csv"),
                 show_col_types = FALSE)
```

## Join everything together

```{r join-data}
# Join sample IDs with metadata (keep only .x versions, rename)
samples <- sample_ids %>%
  left_join(metadata, by = "pid", relationship = "many-to-one") %>%
  rename(arm = arm.x) %>%
  select(-arm.y)

# Widen cytokines (one column per cytokine)
# Clean names: replace hyphens with underscores in ALL column names
luminex_wide <- luminex %>%
  filter(limits == "not_censored") %>%
  select(sample_id, cytokine, conc) %>%
  pivot_wider(names_from = cytokine, values_from = conc)

# Replace hyphens in column names with underscores
names(luminex_wide) <- str_replace_all(names(luminex_wide), "-", "_")

cyt_names <- setdiff(names(luminex_wide), "sample_id")

# Join cytokine values onto samples
data_cyt <- samples %>%
  left_join(luminex_wide, by = "sample_id")

# Join flow cytometry
data_full <- data_cyt %>%
  left_join(flow, by = "sample_id")

glimpse(data_full)
```

---

# 2. Summary Tables with P-values

## 2.1 Cytokine levels by birth control arm

```{r summary-by-arm}
# Helper: run a t-test for each cytokine and return a tidy table
cyt_arm <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, arm)

cyt_arm %>%
  group_by(cytokine, arm) %>%
  summarise(n = n(),
            mean_conc = mean(conc, na.rm = TRUE),
            sd_conc   = sd(conc, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_wider(names_from = arm,
              values_from = c(n, mean_conc, sd_conc),
              names_glue = "{arm}_{.value}") %>%
  # add p-values from t-test
  left_join(
    cyt_arm %>%
      group_by(cytokine) %>%
      summarise(p_value = t.test(conc ~ arm)$p.value, .groups = "drop"),
    by = "cytokine"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kable(caption = "Cytokine concentrations by birth control arm (t-test)")
```

## 2.2 Cytokine levels by time point

```{r summary-by-timepoint}
cyt_time <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, time_point)

cyt_time %>%
  group_by(cytokine, time_point) %>%
  summarise(mean_conc = mean(conc, na.rm = TRUE),
            sd_conc   = sd(conc, na.rm = TRUE),
            n = n(),
            .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kable(caption = "Cytokine concentrations by menstrual cycle time point")
```

```{r anova-timepoint, results = "asis"}
# ANOVA p-values for time point effect on each cytokine
cyt_time %>%
  group_by(cytokine) %>%
  nest() %>%
  mutate(anova_p = map_dbl(data, ~ aov(conc ~ time_point, data = .x) %>%
                             summary() %>% pluck(1, "Pr(>F)", 1))) %>%
  select(cytokine, anova_p) %>%
  mutate(across(where(is.numeric), ~ round(., 4))) %>%
  kable(caption = "ANOVA p-values: does cytokine concentration vary by time point?")
```

## 2.3 PCOS status

```{r summary-pcos}
cyt_pcos <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, pcos_status)

cyt_pcos %>%
  group_by(cytokine, pcos_status) %>%
  summarise(mean_conc = mean(conc, na.rm = TRUE),
            sd_conc   = sd(conc, na.rm = TRUE),
            n = n(),
            .groups = "drop") %>%
  pivot_wider(names_from = pcos_status,
              values_from = c(mean_conc, sd_conc, n),
              names_glue = "{pcos_status}_{.value}") %>%
  left_join(
    cyt_pcos %>%
      group_by(cytokine) %>%
      summarise(p_value = t.test(conc ~ pcos_status)$p.value, .groups = "drop"),
    by = "cytokine"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kable(caption = "Cytokine concentrations by PCOS status (t-test)")
```

## 2.4 Flow cytometry cell counts by birth control arm

```{r flow-summary}
flow_joined <- flow %>%
  left_join(samples, by = "sample_id", relationship = "many-to-one")

# Cell types to test
cell_cols <- c("live_cd19_negative", "cd45_negative", "cd45_positive",
               "neutrophils", "non_neutrophils", "cd3_negative",
               "cd3_positive", "cd4_t_cells", "cd8_t_cells")

flow_joined %>%
  pivot_longer(cols = all_of(cell_cols), names_to = "cell_type",
               values_to = "count") %>%
  group_by(cell_type, arm) %>%
  summarise(mean_count = mean(count, na.rm = TRUE),
            sd_count   = sd(count, na.rm = TRUE),
            n = n(),
            .groups = "drop") %>%
  pivot_wider(names_from = arm,
              values_from = c(mean_count, sd_count, n),
              names_glue = "{arm}_{.value}") %>%
  left_join(
    flow_joined %>%
      pivot_longer(cols = all_of(cell_cols), names_to = "cell_type",
                   values_to = "count") %>%
      group_by(cell_type) %>%
      summarise(p_value = t.test(count ~ arm)$p.value, .groups = "drop"),
    by = "cell_type"
  ) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  kable(caption = "Immune cell counts by birth control arm (t-test)")
```

---

# 3. Graphs

## 3.1 Cytokine boxplots — birth control arm

```{r boxplot-cyt-arm, fig.height = 8}
cyt_long <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, arm)

ggplot(cyt_long, aes(x = arm, y = conc, fill = arm)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ cytokine, scales = "free_y", ncol = 4) +
  scale_y_log10() +
  labs(title = "Cytokine concentrations by birth control arm",
       subtitle = "y-axis on log scale | jitter shows individual samples",
       x = NULL, y = "Concentration (pg/mL)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.2 Cytokine boxplots — time point

```{r boxplot-cyt-time, fig.height = 8}
cyt_long_time <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, time_point)

ggplot(cyt_long_time, aes(x = time_point, y = conc, fill = time_point)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ cytokine, scales = "free_y", ncol = 4) +
  scale_y_log10() +
  labs(title = "Cytokine concentrations across menstrual cycle time points",
       subtitle = "y-axis on log scale",
       x = NULL, y = "Concentration (pg/mL)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.3 Cytokine boxplots — PCOS status

```{r boxplot-cyt-pcos, fig.height = 8}
cyt_long_pcos <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, pcos_status)

ggplot(cyt_long_pcos, aes(x = pcos_status, y = conc, fill = pcos_status)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ cytokine, scales = "free_y", ncol = 4) +
  scale_y_log10() +
  labs(title = "Cytokine concentrations by PCOS status",
       subtitle = "y-axis on log scale",
       x = NULL, y = "Concentration (pg/mL)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.4 Immune cell counts by birth control arm

```{r boxplot-flow-arm, fig.height = 6}
flow_long <- flow_joined %>%
  pivot_longer(cols = all_of(cell_cols), names_to = "cell_type",
               values_to = "count") %>%
  drop_na(count, arm)

ggplot(flow_long, aes(x = arm, y = count, fill = arm)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ cell_type, scales = "free_y", ncol = 3) +
  scale_y_log10() +
  labs(title = "Immune cell counts by birth control arm",
       subtitle = "y-axis on log scale",
       x = NULL, y = "Cell count") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 3.5 Correlation heatmap — cytokines and immune cells

```{r corr-heatmap, fig.height = 6}
library(corrplot)

# Build a wide dataset with cytokines + key immune cells
corr_data <- data_full %>%
  select(all_of(cell_cols), all_of(cyt_names)) %>%
  drop_na()

cm <- cor(corr_data, use = "pairwise.complete.obs")

corrplot(cm, method = "color", type = "upper",
         tl.col = "black", tl.cex = 0.7,
         addCoef.col = "black", number.cex = 0.5,
         title = "Correlation: Cytokines × Immune Cells",
         mar = c(0, 0, 2, 0))
```

## 3.6 Scatter plot — strongest cytokine–immune cell relationship

```{r scatter-example, fig.height = 5}
data_full %>%
  drop_na(IL_6, neutrophils) %>%
  ggplot(aes(x = IL_6, y = neutrophils, color = arm)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "IL-6 vs Neutrophil count",
       subtitle = "Coloured by birth control arm | both axes on log scale",
       x = "IL-6 concentration (pg/mL)",
       y = "Neutrophil count") +
  theme_minimal()
```

---

# 4. Regression Models

## 4.1 Linear regression: cytokine ~ arm + age + pcos_status

We model each cytokine as a function of birth control arm, age, and PCOS status.

```{r reg-cytokines, results = "asis"}
cyt_reg <- data_cyt %>%
  pivot_longer(cols = all_of(cyt_names),
               names_to = "cytokine", values_to = "conc") %>%
  drop_na(conc, arm, age, pcos_status) %>%
  mutate(conc_log = log10(conc + 1))

models <- cyt_reg %>%
  group_by(cytokine) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(conc_log ~ arm + age + pcos_status, data = .x)),
    tidied = map(model, tidy)
  ) %>%
  unnest(tidied) %>%
  select(cytokine, term, estimate, std.error, statistic, p.value) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

kable(models, caption = "Linear regression: log10(cytokine) ~ arm + age + pcos_status")
```

**Interpretation:** The `arm` coefficient tells us the difference in
log-cytokine concentration between the `birth_control` and
`no_birth_control` groups, holding age and PCOS status constant. A
significant p-value (< 0.05) suggests the birth control arm has a
meaningful association with that cytokine.

## 4.2 Linear regression: immune cells ~ arm + age + pcos_status

```{r reg-flow, results = "asis"}
flow_reg <- flow_joined %>%
  pivot_longer(cols = all_of(cell_cols), names_to = "cell_type",
               values_to = "count") %>%
  drop_na(count, arm, age, pcos_status) %>%
  mutate(count_log = log10(count + 1))

models_flow <- flow_reg %>%
  group_by(cell_type) %>%
  nest() %>%
  mutate(
    model = map(data, ~ lm(count_log ~ arm + age + pcos_status, data = .x)),
    tidied = map(model, tidy)
  ) %>%
  unnest(tidied) %>%
  select(cell_type, term, estimate, std.error, statistic, p.value) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

kable(models_flow,
      caption = "Linear regression: log10(cell count) ~ arm + age + pcos_status")
```

---

# 5. Quick Takeaways

```{r takeaways, results = "asis"}
# Count significant arm effects in cytokine regression
sig_cyt <- models %>%
  filter(term == "armbirth_control", p.value < 0.05) %>%
  pull(cytokine)

sig_flow <- models_flow %>%
  filter(term == "armbirth_control", p.value < 0.05) %>%
  pull(cell_type)

cat("**Cytokines significantly associated with birth control (p < 0.05):**  \n")
if (length(sig_cyt) > 0) {
  cat(paste0("- ", sig_cyt, collapse = "\n"))
} else {
  cat("- None (no significant associations detected)")
}

cat("\n\n**Immune cell types significantly associated with birth control (p < 0.05):**  \n")
if (length(sig_flow) > 0) {
  cat(paste0("- ", sig_flow, collapse = "\n"))
} else {
  cat("- None (no significant associations detected)")
}
```