---
title: "Bibliometric Co-occurrence Network — Sarcoptes scabiei Diagnostics"
subtitle: "Built from diagnostic accuracy study data"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  message: false
  warning: false
keep-tex: true
---

## Overview

This network visualises co-occurrence relationships in the diagnostic
accuracy literature on scabies (*Sarcoptes scabiei*). Each edge represents a
study that links S. scabiei to a host, a diagnostic method, a sample type,
or an antigen type. The thicker the edge, the more studies support that link.

```{r setup}
library(tidyverse)
library(janitor)

raw <- read_csv("../data/Diagnostic_Accuracy_Table.csv", show_col_types = FALSE)

# Clean data: remove non-ASCII, filter valid rows
d <- raw |>
  clean_names() |>
  filter(!is.na(tp)) |>
  mutate(across(where(is.character), ~ iconv(., from = "latin1", to = "ASCII//TRANSLIT"))) |>
  mutate(across(where(is.character), ~ gsub("[^ -~]", "", .)))

# --- Categorise variables for the network ---

d <- d |> mutate(
  host_group = case_when(
    tolower(study_on) == "human" ~ "Humans",
    TRUE ~ "Animals"
  ),
  method_group = case_when(
    method %in% c("qPCR", "Qpcr", "nqPCR", "nPCR", "cPCR") ~ "qPCR / cPCR",
    method == "LAMP" ~ "LAMP",
    method == "nRT-qPCR" ~ "RT-qPCR",
    grepl("ELISA|DELFIA", method, ignore.case = TRUE) ~ "ELISA"
  ),
  sample_group = case_when(
    tolower(sample_type) == "serum" ~ "Serum",
    grepl("scrap", tolower(sample_type)) ~ "Skin scraping",
    grepl("swab", tolower(sample_type)) ~ "Swab",
    TRUE ~ sample_type
  ),
  antigen_group = case_when(
    grepl("wma", tolower(antigen_type)) ~ "Whole antigen",
    grepl("recombinant", tolower(antigen_type)) ~ "Recombinant",
    tolower(antigen_type) %in% c("mtdna", "nuclear genes", "microsatellite marker") ~ "DNA target"
  )
)

# --- Build co-occurrence edge weights ---

# Edge: Sarcoptes ↔ host
host_edges <- d |> count(host_group, name = "weight") |> mutate(to = host_group, from = "Sarcoptes scabiei")
# Edge: Sarcoptes ↔ method
method_edges <- d |> filter(!is.na(method_group)) |> count(method_group, name = "weight") |> mutate(to = method_group, from = "Sarcoptes scabiei")
# Edge: Sarcoptes ↔ sample type
sample_edges <- d |> count(sample_group, name = "weight") |> mutate(to = sample_group, from = "Sarcoptes scabiei")
# Edge: Sarcoptes ↔ antigen type
antigen_edges <- d |> filter(!is.na(antigen_group)) |> count(antigen_group, name = "weight") |> mutate(to = antigen_group, from = "Sarcoptes scabiei")

# Edge: Serum ↔ ELISA (all 40 ELISA studies used serum, but let's count)
serum_elisa <- d |> filter(method_group == "ELISA", sample_group == "Serum") |> nrow()

all_edges <- bind_rows(host_edges, method_edges, sample_edges, antigen_edges)
all_edges <- all_edges |> filter(!is.na(to))
```

---

## Co-occurrence network

```{r}
#| fig-height: 5.5
#| fig-width: 7

# ---- Layout: assign (x, y) positions for each node ----

# Outer ring: node groups with their positions
all_nodes <- c("Sarcoptes scabiei",
  "Humans", "Animals",
  "ELISA", "qPCR / cPCR", "LAMP", "RT-qPCR",
  "Serum", "Skin scraping", "Swab", "meat-juice",
  "Whole antigen", "Recombinant", "DNA target")

# Positions — compact, hub at centre
node_pos <- list(
  "Sarcoptes scabiei" = c(0, 0),
  "Humans" = c(-1.4, -0.4),
  "Animals" = c(-1.4, -1.3),
  "ELISA" = c(0.6, 1.7),
  "qPCR / cPCR" = c(-0.6, 1.4),
  "LAMP" = c(-1.1, 0.7),
  "RT-qPCR" = c(0.0, 0.8),
  "Serum" = c(1.7, 1.0),
  "Skin scraping" = c(1.5, 0.0),
  "Swab" = c(1.7, -0.7),
  "meat-juice" = c(1.5, -1.3),
  "Whole antigen" = c(-0.3, -1.3),
  "Recombinant" = c(0.4, -1.0),
  "DNA target" = c(1.0, -1.4)
)

# Colours for groups
group_colors <- c(
  "Sarcoptes scabiei" = "#E67E22",
  "Humans" = "#1ABC9C",
  "Animals" = "#8B4513",
  "ELISA" = "#2ECC71",
  "qPCR / cPCR" = "#3498DB",
  "LAMP" = "#2980B9",
  "RT-qPCR" = "#5DADE2",
  "Serum" = "#27AE60",
  "Skin scraping" = "#C0392B",
  "Swab" = "#E74C3C",
  "meat-juice" = "#A93226",
  "Whole antigen" = "#D4AC0D",
  "Recombinant" = "#F39C12",
  "DNA target" = "#8E44AD"
)

# Labels for groups (display names)
display_name <- c(
  "Sarcoptes scabiei" = "Sarcoptes scabiei",
  "Humans" = "Humans",
  "Animals" = "Animals",
  "ELISA" = "ELISA",
  "qPCR / cPCR" = "qPCR/cPCR",
  "LAMP" = "LAMP",
  "RT-qPCR" = "RT-qPCR",
  "Serum" = "Serum",
  "Skin scraping" = "Skin scraping",
  "Swab" = "Swab",
  "meat-juice" = "Meat juice",
  "Whole antigen" = "Whole antigen",
  "Recombinant" = "Recombinant",
  "DNA target" = "DNA target"
)

# ---- Plot ----

# Set up blank canvas
par(mar = c(0, 0, 0, 0), bg = "#FAFAFA")
plot(NA, xlim = c(-2.5, 2.5), ylim = c(-2.0, 2.3),
     xlab = "", ylab = "", xaxt = "n", yaxt = "n", bty = "n")

# Title
text(0, 2.1, "Bibliometric Co-occurrence Network", cex = 0.9, font = 2, xpd = TRUE)
text(0, 1.9, "Sarcoptes scabiei Diagnostics", cex = 0.65, font = 3, col = "#666666", xpd = TRUE)

# Draw edges: Sarcoptes ↔ each other node
# Edge width proportional to sqrt(weight) for visual clarity
for (i in seq_len(nrow(all_edges))) {
  e <- all_edges[i, ]
  x0 <- node_pos[[e$from]][1]
  y0 <- node_pos[[e$from]][2]
  x1 <- node_pos[[e$to]][1]
  y1 <- node_pos[[e$to]][2]

  if (!is.null(x0) && !is.null(x1)) {
    w <- sqrt(e$weight) * 0.6
    segments(x0, y0, x1, y1,
             col = adjustcolor(group_colors[e$to], alpha.f = 0.5),
             lwd = w)
  }
}

# Special edge: Serum → ELISA (moderator relationship)
x0 <- node_pos[["Serum"]][1]
y0 <- node_pos[["Serum"]][2]
x1 <- node_pos[["ELISA"]][1]
y1 <- node_pos[["ELISA"]][2]
segments(x0, y0, x1, y1,
         col = adjustcolor("#27AE60", alpha.f = 0.5), lwd = sqrt(serum_elisa) * 0.8, lty = 2)

# Draw nodes — scaled down
for (node in all_nodes) {
  pos <- node_pos[[node]]
  if (!is.null(pos)) {
    col <- group_colors[node]
    # Base circle size: hub bigger
    r <- ifelse(node == "Sarcoptes scabiei", 0.45, 0.28)
    # Outer glow
    symbols(pos[1], pos[2], circles = r * 1.3,
            inches = FALSE, add = TRUE,
            bg = adjustcolor(col, alpha.f = 0.25),
            fg = NA)
    # Main node
    symbols(pos[1], pos[2], circles = r,
            inches = FALSE, add = TRUE,
            bg = col,
            fg = "white", lwd = 1.5)
    # Label
    lbl <- display_name[node]
    cex_lbl <- ifelse(node == "Sarcoptes scabiei", 0.55, 0.45)
    font_lbl <- ifelse(node == "Sarcoptes scabiei", 2, 1)
    text(pos[1], pos[2] - r - 0.08,
         lbl, cex = cex_lbl, font = font_lbl,
         col = "#222222", xpd = TRUE)
  }
}

# Legend — smaller
legend("topright",
  legend = c(
    "Sarcoptes scabiei (hub)",
    "Host: Humans", "Host: Animals",
    "Molecular diagnostics", "Immunological diagnostics",
    "Serum (moderator)", "Clinical / morphological"
  ),
  fill = c("#E67E22", "#1ABC9C", "#8B4513",
           "#3498DB", "#2ECC71",
           "#27AE60", "#C0392B"),
  cex = 0.5, bty = "n", title = "", title.adj = 0,
  xpd = TRUE
)

# Note
mtext("Edge thickness = number of studies", side = 1, line = -1,
      cex = 0.5, col = "#888888", font = 3, adj = 1)
```

---

## Edge weights table

```{r}
all_edges |>
  select(from, to, weight) |>
  arrange(desc(weight)) |>
  knitr::kable(caption = "Co-occurrence edge weights (number of studies)")
```

```{r}
cat("ELISA studies using Serum:", serum_elisa, "out of",
    sum(d$method_group == "ELISA", na.rm = TRUE), "total ELISA studies")
```

---

## Interpretation

- **Sarcoptes scabiei** (orange, centre) connects to all other nodes.
- **Hosts:** 16 studies on humans (teal), 37 on animals (brown — pigs,
  rabbits, dogs, foxes, sheep, chamois, bears).
- **Molecular diagnostics** (blues): qPCR/cPCR (8 studies), LAMP (1),
  RT-qPCR (2).
- **Immunological diagnostics** (green): ELISA dominates with 42 studies.
  Nearly all ELISA studies used **serum** (dashed green edge) — serum acts as
  a *moderator* variable between the diagnostic method and the pathogen.
- **Clinical/morphological** (reds): Skin scrapings and swabs are the sample
  collection methods used across both diagnostic approaches.
- **Antigen types** (golds/purple): Whole antigen (16), recombinant (23),
  DNA target (11).
- Edge thickness reflects the number of linking studies — more evidence
  behind that connection.