---
title: "Diagnostic Accuracy Meta-Analysis"
subtitle: "Sensitivity, Specificity, ROC, Publication Bias & More"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
  pdf:
    toc: true
    keep-tex: true
execute:
  message: false
  warning: false
---

## Overview

This document performs a full diagnostic accuracy meta-analysis on the data
in `data/Diagnostic_Accuracy_Table.csv` using the `mada` and `meta` packages.

We'll cover:

1.  Load & clean data
2.  Sensitivity / Specificity with heterogeneity (I²)
3.  Subgroup analysis (by diagnostic type)
4.  Diagnostic odds ratio (DOR)
5.  Forest plot
6.  SROC curve
7.  Funnel plot & publication bias test

```{r setup}
library(tidyverse)
library(mada)
library(meta)
```

---

## 1. Load & prepare the data

```{r}
raw <- read_csv("../data/Diagnostic_Accuracy_Table.csv", show_col_types = FALSE)

# Clean column names and filter out rows with missing TP/FP/FN/TN
diag <- raw |>
  janitor::clean_names() |>
  filter(!is.na(tp), !is.na(fp), !is.na(tn), !is.na(fn)) |>
  # Strip non-ASCII characters that break plotly hover text
  mutate(across(where(is.character), ~ iconv(., from = "latin1", to = "ASCII//TRANSLIT"))) |>
  mutate(across(where(is.character), ~ gsub("[^ -~]", "", .))) |>
  # Add study-level proportions for mada functions
  mutate(
    study_label = str_glue("{study} ({country})"),
    n = tp + fp + tn + fn,
    prev = (tp + fn) / n       # disease prevalence in each study
  ) |>
  # mada::reitsma() requires exact uppercase TP, FN, FP, TN column names
  rename(TP = tp, FN = fn, FP = fp, TN = tn) |>
  # Remove % signs from sensitivity/specificity columns
  mutate(
    sensitivity = as.numeric(gsub("%", "", sensitivity)),
    specificity = as.numeric(gsub("%", "", specificity))
  )

cat(nrow(diag), "studies loaded\n")

# Quick peek
diag |>
  select(study_label, TP, FP, TN, FN, diagnostic_type) |>
  head(10) |>
  knitr::kable()
```

---

## 2. Sensitivity, Specificity & Heterogeneity

Calculate pooled sensitivity and specificity with a bivariate random-effects
model (the Reitsma model in `mada`). This accounts for the negative
correlation between sensitivity and specificity.

```{r}
# Fit the bivariate model
fit <- mada::reitsma(diag, method = "reml")
summary(fit)

# Extract the summary table
s <- summary(fit)
```

```{r}
# Pooled estimates from the model
# s$coefficients has logit row 1-2 and back-transformed row 3-4
sens_lb <- s$coefficients["sensitivity", "95%ci.lb"]
sens_ub <- s$coefficients["sensitivity", "95%ci.ub"]
sens <- s$coefficients["sensitivity", "Estimate"]

fpr_lb <- s$coefficients["false pos. rate", "95%ci.lb"]
fpr_ub <- s$coefficients["false pos. rate", "95%ci.ub"]
fpr <- s$coefficients["false pos. rate", "Estimate"]

# Specificity = 1 − FPR
spec_lb <- 1 - fpr_ub
spec_ub <- 1 - fpr_lb
spec <- 1 - fpr

# DOR from logit coefficients
logit_sens <- s$coefficients[1, "Estimate"]
logit_fpr <- s$coefficients[2, "Estimate"]
dor <- exp(logit_sens - logit_fpr)
dor_lb <- exp(s$coefficients[1, "95%ci.lb"] - s$coefficients[2, "95%ci.ub"])
dor_ub <- exp(s$coefficients[1, "95%ci.ub"] - s$coefficients[2, "95%ci.lb"])

pooled <- data.frame(
  Measure = c("Sensitivity", "Specificity", "DOR"),
  Estimate = c(
    round(sens * 100, 1),
    round(spec * 100, 1),
    round(dor, 2)
  ),
  CI_lower = c(
    round(sens_lb * 100, 1),
    round(spec_lb * 100, 1),
    round(dor_lb, 2)
  ),
  CI_upper = c(
    round(sens_ub * 100, 1),
    round(spec_ub * 100, 1),
    round(dor_ub, 2)
  )
)

pooled |>
  knitr::kable(caption = "Pooled estimates (bivariate random-effects model)")
```

### Heterogeneity (I²)

```{r}
# We compute I² from univariate meta-analyses on logit(sens) and logit(spec)
# to get a feel for heterogeneity

diag_for_meta <- diag |>
  mutate(
    logit_sens = log(TP / (TP + FN + 0.5)),
    se_logit_sens = sqrt(1 / (TP + 0.5) + 1 / (FN + 0.5)),
    logit_spec = log(TN / (TN + FP + 0.5)),
    se_logit_spec = sqrt(1 / (TN + 0.5) + 1 / (FP + 0.5))
  )

meta_sens <- metagen(
  TE = logit_sens, seTE = se_logit_sens,
  data = diag_for_meta,
  studlab = study_label,
  sm = "PLOGIT",
  fixed = FALSE, random = TRUE,
  title = "Sensitivity"
)

meta_spec <- metagen(
  TE = logit_spec, seTE = se_logit_spec,
  data = diag_for_meta,
  studlab = study_label,
  sm = "PLOGIT",
  fixed = FALSE, random = TRUE,
  title = "Specificity"
)

cat("=== Heterogeneity ===\n\n")
cat("Sensitivity: I² =", round(meta_sens$I2, 1), "%  ",
    "τ² =", round(meta_sens$tau2, 3), "\n")
cat("Specificity: I² =", round(meta_spec$I2, 1), "%  ",
    "τ² =", round(meta_spec$tau2, 3), "\n")
```

---

## 3. Subgroup analysis — by diagnostic type

The main distinction in the data is `Immunological` (ELISA-based) vs.
`Molecular` (PCR-based). Let's compare them.

```{r}
diag |>
  count(diagnostic_type) |>
  knitr::kable(caption = "Studies by diagnostic type")
```

```{r}
# Fit bivariate model separately for each subgroup
subgroups <- split(diag, diag$diagnostic_type)

results_list <- list()
for (nm in names(subgroups)) {
  if (nrow(subgroups[[nm]]) >= 4) {
    f <- mada::reitsma(subgroups[[nm]], method = "reml")
    s <- summary(f)
    # Back-transformed values are in rows 3-4
    sens_val <- s$coefficients["sensitivity", "Estimate"]
    sens_lb <- s$coefficients["sensitivity", "95%ci.lb"]
    sens_ub <- s$coefficients["sensitivity", "95%ci.ub"]
    fpr_val <- s$coefficients["false pos. rate", "Estimate"]
    fpr_lb <- s$coefficients["false pos. rate", "95%ci.lb"]
    fpr_ub <- s$coefficients["false pos. rate", "95%ci.ub"]
    spec <- 1 - fpr_val
    spec_lb <- 1 - fpr_ub
    spec_ub <- 1 - fpr_lb
    # DOR from logit coefficients
    logit_sens <- s$coefficients[1, "Estimate"]
    logit_fpr <- s$coefficients[2, "Estimate"]
    dor_val <- exp(logit_sens - logit_fpr)
    dor_lb <- exp(s$coefficients[1, "95%ci.lb"] - s$coefficients[2, "95%ci.ub"])
    dor_ub <- exp(s$coefficients[1, "95%ci.ub"] - s$coefficients[2, "95%ci.lb"])
    results_list[[nm]] <- data.frame(
      Subgroup = nm,
      n_studies = nrow(subgroups[[nm]]),
      Sensitivity = paste0(round(sens_val * 100, 1), "% [",
                           round(sens_lb * 100, 1), "% – ",
                           round(sens_ub * 100, 1), "%]"),
      Specificity = paste0(round(spec * 100, 1), "% [",
                           round(spec_lb * 100, 1), "% – ",
                           round(spec_ub * 100, 1), "%]"),
      DOR = paste0(round(dor_val, 2), " [",
                   round(dor_lb, 2), " – ",
                   round(dor_ub, 2), "]")
    )
  }
}

bind_rows(results_list) |>
  knitr::kable(caption = "Subgroup analysis by diagnostic type")
```

```{r}
# Visual comparison on the ROC plane
plot(x = 1, type = "n", xlim = c(0, 1), ylim = c(0, 1),
     xlab = "1 − Specificity (FPR)", ylab = "Sensitivity (TPR)",
     main = "Subgroup comparison on the ROC plane")
abline(a = 0, b = 1, lty = 2, col = "grey80")
colors <- c("Immunological" = "#2166ac", "Molecular" = "#b2182b")
for (nm in names(subgroups)) {
  if (nrow(subgroups[[nm]]) >= 4) {
    f <- mada::reitsma(subgroups[[nm]], method = "reml")
    mada::ROCellipse(f, col = colors[nm], lwd = 2)
    # Pooled point
    s <- summary(f)
    sens <- plogis(s$coefficients[1, 1])
    spec <- plogis(s$coefficients[2, 1])
    points(1 - spec, sens, pch = 19, col = colors[nm], cex = 1.5)
  }
}
legend("bottomright", legend = names(colors), fill = colors, title = "Diagnostic type")
```

---

<!-- markdownlint-disable MD033 -->
## 4. Forest plot — Sensitivity & Specificity

```{r}
#| fig-height: 12
#| fig-width: 10

# Forest plot using meta package on logit-transformed proportions
# Back-transform to proportions for the plot

forest(meta_sens,
  leftcols = c("studlab"),
  rightcols = c("effect", "ci", "w.random"),
  leftlabs = c("Study"),
  rightlabs = c("Sens.", "95% CI", "Weight"),
  smlab = "Sensitivity (logit scale → back-transformed)",
  pooled.totals = TRUE,
  hetstat = TRUE,
  overall = TRUE,
  digits = 2
)
```

```{r}
#| fig-height: 12
#| fig-width: 10

forest(meta_spec,
  leftcols = c("studlab"),
  rightcols = c("effect", "ci", "w.random"),
  leftlabs = c("Study"),
  rightlabs = c("Spec.", "95% CI", "Weight"),
  smlab = "Specificity (logit scale → back-transformed)",
  pooled.totals = TRUE,
  hetstat = TRUE,
  overall = TRUE,
  digits = 2
)
```

---

## 5. Diagnostic Odds Ratio (DOR) forest plot

The DOR = (TP/FP) / (FN/TN) — a single-number summary of test performance.
DOR > 1 means the test discriminates diseased from non-diseased.

```{r}
# Calculate DOR per study (with continuity correction for zeros)
diag_for_dor <- diag |>
  mutate(
    tp_c = TP + 0.5,
    fp_c = FP + 0.5,
    fn_c = FN + 0.5,
    tn_c = TN + 0.5,
    log_dor = log((tp_c / fp_c) / (fn_c / tn_c)),
    se_log_dor = sqrt(1 / tp_c + 1 / fp_c + 1 / fn_c + 1 / tn_c)
  )

meta_dor <- metagen(
  TE = log_dor, seTE = se_log_dor,
  data = diag_for_dor,
  studlab = study_label,
  sm = "OR",
  fixed = FALSE, random = TRUE,
  title = "Diagnostic Odds Ratio"
)
```

```{r}
#| fig-height: 12
#| fig-width: 10

forest(meta_dor,
  leftcols = c("studlab"),
  rightcols = c("effect", "ci", "w.random"),
  leftlabs = c("Study"),
  rightlabs = c("DOR", "95% CI", "Weight"),
  smlab = "Diagnostic Odds Ratio",
  pooled.totals = TRUE,
  hetstat = TRUE,
  overall = TRUE,
  digits = 2
)
```

```{r}
cat("Pooled DOR =", round(exp(meta_dor$TE.random), 2),
    "95% CI:", round(exp(meta_dor$lower.random), 2), "–",
    round(exp(meta_dor$upper.random), 2), "\n")
cat("Heterogeneity: I² =", round(meta_dor$I2, 1), "%\n")
```

---

## 6. SROC Curve (Summary ROC)

The bivariate model can be plotted as an SROC curve. This shows the
trade-off between sensitivity and specificity across all studies.

```{r}
#| fig-height: 6
#| fig-width: 7

# Build the SROC plot manually using mada's sroc() and ROCellipse()

# Start with an empty plot
plot(NA, xlim = c(0, 1), ylim = c(0, 1),
     xlab = "False Positive Rate (1 − Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     main = "SROC Curve with 95% confidence & prediction regions")
abline(a = 0, b = 1, lty = 2, col = "grey80")

# Prediction region (larger ellipse)
mada::ROCellipse(fit, pch = NA, lty = 3, lwd = 1.5)

# Confidence region (smaller ellipse)
mada::ROCellipse(fit, pch = NA, level = 0.95,
                 lty = 2, lwd = 2, col = "black")

# SROC curve
fpr_seq <- seq(0.01, 0.99, 0.01)
sroc_curve <- mada::sroc(fit, fpr = fpr_seq)
lines(sroc_curve, lwd = 2, col = "black")

# Pooled estimate point
sens_pooled <- s$coefficients["sensitivity", "Estimate"]
fpr_pooled <- s$coefficients["false pos. rate", "Estimate"]
points(fpr_pooled, sens_pooled, pch = 19, cex = 1.5)

# Individual study points
text(1 - diag$specificity / 100,
     diag$sensitivity / 100,
     labels = seq_len(nrow(diag)),
     cex = 0.6, col = "grey40")

legend("bottomright",
  legend = c("Observed study", "SROC curve",
             "95% confidence region", "95% prediction region"),
  pch = c(1, NA, NA, NA),
  lty = c(NA, 1, 2, 3),
  col = c("black", "black", "black", "black"),
  cex = 0.8
)
```

---

## 7. Interactive funnel plot — Diagnostic Odds Ratio

We use the DOR (log scale) for the funnel plot because it's a single
univariate measure. Asymmetry may indicate publication bias. Hover over
each point to see the study details.

```{r}
#| fig-height: 6
#| fig-width: 8
#| eval: !expr knitr::is_html_output()

library(plotly)

# Build a data frame with the funnel plot variables
funnel_data <- data.frame(
  study = diag_for_dor$study_label,
  log_dor = meta_dor$TE,
  se = meta_dor$seTE,
  dor = exp(meta_dor$TE),
  diagnostic_type = diag_for_dor$diagnostic_type,
  country = diag_for_dor$country,
  study_on = diag_for_dor$study_on,
  method = diag_for_dor$method
) |>
  mutate(
    # Create hover text
    hover = paste0(
      "<b>", study, "</b><br>",
      "DOR: ", round(dor, 1), "<br>",
      "log(DOR): ", round(log_dor, 2), "<br>",
      "SE: ", round(se, 3), "<br>",
      "Type: ", diagnostic_type, "<br>",
      "Host: ", study_on, "<br>",
      "Country: ", country
    )
  )

# Pseudo95% confidence limits around the pooled estimate
pooled_log_dor <- meta_dor$TE.random
y_lim <- max(funnel_data$se) * 1.15
x_range <- max(abs(funnel_data$log_dor - pooled_log_dor)) * 1.3
se_seq <- seq(0, y_lim, length.out = 50)

# Funnel triangle boundaries (pseudo-95% CI)
ci_upper <- pooled_log_dor + 1.96 * se_seq
ci_lower <- pooled_log_dor - 1.96 * se_seq

fig <- plot_ly()

# Add the pseudo-95% CI lines
fig <- fig |> add_trace(
  x = ~ci_upper, y = ~se_seq,
  type = "scatter", mode = "lines",
  line = list(color = "grey60", dash = "dash"),
  showlegend = FALSE,
  hoverinfo = "none"
)
fig <- fig |> add_trace(
  x = ~ci_lower, y = ~se_seq,
  type = "scatter", mode = "lines",
  line = list(color = "grey60", dash = "dash"),
  showlegend = FALSE,
  hoverinfo = "none"
)

# Vertical line at pooled estimate
fig <- fig |> add_trace(
  x = c(pooled_log_dor, pooled_log_dor), y = c(0, y_lim),
  type = "scatter", mode = "lines",
  line = list(color = "grey40", dash = "dot"),
  showlegend = FALSE,
  hoverinfo = "none"
)

# Add the study points, coloured by diagnostic type
colors <- c("Immunological" = "#228B22", "Molecular" = "#4169E1")
for (dt in unique(funnel_data$diagnostic_type)) {
  sub <- funnel_data[funnel_data$diagnostic_type == dt, ]
  fig <- fig |> add_trace(
    x = ~log_dor, y = ~se,
    data = sub,
    type = "scatter", mode = "markers",
    marker = list(
      size = 8,
      color = colors[dt],
      line = list(color = "black", width = 0.5)
    ),
    text = ~hover,
    hoverinfo = "text",
    name = dt
  )
}

fig <- fig |> layout(
  title = "Interactive funnel plot — Diagnostic Odds Ratio",
  xaxis = list(
    title = "log(Diagnostic Odds Ratio)",
    zeroline = FALSE,
    range = c(pooled_log_dor - x_range, pooled_log_dor + x_range)
  ),
  yaxis = list(
    title = "Standard Error",
    autorange = "reversed",
    zeroline = FALSE,
    range = c(y_lim, 0)
  ),
  hovermode = "closest",
  legend = list(title = list(text = "<b>Diagnostic type</b>"))
)

fig
```

```{r}
#| eval: !expr knitr::is_latex_output()
#| fig-height: 6
#| fig-width: 6

funnel(meta_dor,
  studlab = TRUE,
  xlab = "Diagnostic Odds Ratio (log scale)",
  ylab = "Standard Error",
  main = "Funnel plot — Diagnostic Odds Ratio"
)
```

### Egger's test for funnel asymmetry

```{r}
# Egger's test: regress standard normal deviate on precision
# A significant intercept (p < 0.10) suggests asymmetry
egger <- metabias(meta_dor, method.bias = "linreg",
                  k.min = 10)  # minimum studies to test
print(egger)

cat("\nEgger's test p-value:", round(egger$pval, 4))
if (egger$pval < 0.10) {
  cat(" → Evidence of asymmetry (possible publication bias)")
} else {
  cat(" → No strong evidence of asymmetry")
}
```

---

## Summary table

```{r}
pooled |>
  knitr::kable(caption = "Final pooled estimates")
```

Key takeaways:

-   **Sensitivity (pooled):** `r pooled$Estimate[1]`% (95% CI: `r pooled$CI_lower[1]`% – `r pooled$CI_upper[1]`%)
-   **Specificity (pooled):** `r pooled$Estimate[2]`% (95% CI: `r pooled$CI_lower[2]`% – `r pooled$CI_upper[2]`%)
-   **DOR (pooled):** `r pooled$Estimate[3]` (95% CI: `r pooled$CI_lower[3]` – `r pooled$CI_upper[3]`)
-   Heterogeneity: sensitivity I² = `r round(meta_sens$I2, 1)`%, specificity I² = `r round(meta_spec$I2, 1)`%
-   Egger's test: p = `r round(egger$pval, 4)`