---
title: "Immune cell & cytokine dynamics across the menstrual cycle"
subtitle: "Effects of hormonal birth control and PCOS status on vaginal immune populations"
author: "Group B — Menstruation Study"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  message: false
  warning: false
---



## Research question

How do immune cell populations and inflammatory cytokines change across the
menstrual cycle, and are these patterns affected by hormonal birth control
and/or PCOS status?

## Study design

- **Longitudinal study:** 27 female participants, 4 time points across the
  menstrual cycle (week_prior → onset → end_bleeding → week_post)
- **108 vaginal swab samples** total
- **14** participants use hormonal birth control, **13** do not
- **~50%** have PCOS (polycystic ovary syndrome)
- **Two assays:**
  - Flow cytometry → immune cell counts (CD45+, neutrophils, T cells)
  - Luminex → 8 inflammatory cytokines (IL-1a, IL-1b, IL-6, TNFa, etc.)

```{r}
#| include: false
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(car)
library(gt)
library(ggpubr)

theme_set(theme_minimal(base_size = 11))

# Load data
sample_ids <- read_csv("../data/group_b_menstruation/00_sample_ids_period.csv")
metadata   <- read_csv("../data/group_b_menstruation/01_participant_metadata_period.csv")
luminex    <- read_csv("../data/group_b_menstruation/02_luminex_period.csv")
flow       <- read_csv("../data/group_b_menstruation/03_flow_cytometry_period.csv")

# Master dataset
master <- flow |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  mutate(
    time_point = fct_relevel(time_point,
      "week_prior", "onset", "end_bleeding", "week_post"),
    cd4_cd8_ratio = cd4_t_cells / cd8_t_cells
  )

# Luminex master
luminex_master <- luminex |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  mutate(time_point = fct_relevel(time_point,
    "week_prior", "onset", "end_bleeding", "week_post"))
```

---

## Paper-quality tables

### Table 1 — Immune cell counts across the cycle

```{r}
#| fig-cap: "Table 1"
desc_table <- master |>
  group_by(time_point, arm) |>
  summarise(
    n = n(),
    `CD45+`   = sprintf("%.0f (%.0f)", median(cd45_positive), IQR(cd45_positive)),
    Neutrophils = sprintf("%.0f (%.0f)", median(neutrophils), IQR(neutrophils)),
    `CD4+ T`  = sprintf("%.0f (%.0f)", median(cd4_t_cells), IQR(cd4_t_cells)),
    `CD8+ T`  = sprintf("%.0f (%.0f)", median(cd8_t_cells), IQR(cd8_t_cells)),
    `CD4/CD8` = sprintf("%.2f (%.2f)", median(cd4_cd8_ratio, na.rm = TRUE),
                           IQR(cd4_cd8_ratio, na.rm = TRUE)),
    .groups = "drop"
  )

desc_table |>
  gt(groupname_col = "time_point") |>
  tab_header(
    title = md("**Table 1.** Immune cell counts across the menstrual cycle"),
    subtitle = "Median (IQR) by time point and study arm. n = 108 samples from 27 participants."
  ) |>
  fmt_markdown(columns = everything()) |>
  cols_label(
    arm = "Study arm",
    n = "n",
    `CD45+` = "CD45+",
    `Neutrophils` = "Neutrophils",
    `CD4+ T` = "CD4+ T cells",
    `CD8+ T` = "CD8+ T cells",
    `CD4/CD8` = "CD4/CD8"
  ) |>
  tab_spanner(label = "Cell counts (cells / sample)", columns = 3:8) |>
  tab_source_note(source_note = "Birth control arm: n = 14 participants. No birth control arm: n = 13.") |>
  tab_options(
    heading.title.font.size = px(14),
    column_labels.font.size = px(11),
    table.font.size = px(10)
  )
```

### Table 2 — Cytokine concentrations

```{r}
#| fig-cap: "Table 2"
cytokines_keep <- c("IL-1a", "IL-6", "IP-10", "MIP-3a")
luminex_master |>
  filter(limits == "not_censored", cytokine %in% cytokines_keep) |>
  group_by(time_point, arm, cytokine) |>
  summarise(
    n = n(),
    `Median (IQR) pg/mL` = sprintf("%.2f (%.2f)", median(conc, na.rm = TRUE),
                                     IQR(conc, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  gt(groupname_col = "cytokine") |>
  tab_header(
    title = md("**Table 2.** Cytokine concentrations across the menstrual cycle"),
    subtitle = "Median (IQR) pg/mL; only 'not censored' values shown."
  ) |>
  cols_label(
    time_point = "Time point",
    arm = "Study arm",
    n = "n"
  ) |>
  tab_source_note(source_note = "IL-1a and IP-10 had the highest detection rates. TNFa and IFNg were frequently below detection.")
```

---

## Regression models

We fit **linear mixed-effects models** (random intercept per participant) with
log-transformed outcomes. Type III ANOVA with Satterthwaite denominator df.

### Model 1: CD45+ immune cells ~ time point × arm

```{r}
m1 <- lmer(log(cd45_positive) ~ time_point * arm + (1 | pid), data = master)
anova(m1, type = 3) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  gt() |>
  tab_header(
    title = md("**Model 1.** log(CD45+) ~ time_point × arm"),
    subtitle = "Type III ANOVA (Satterthwaite)"
  ) |>
  fmt_number(columns = c("Sum Sq", "Mean Sq"), decimals = 3) |>
  fmt_number(columns = c("F value"), decimals = 2) |>
  fmt_number(columns = c("Pr(>F)"), decimals = 4)
```

**Result:** No significant main effect of time point (F~3,75~ = 1.67, p = 0.18),
arm (F~1,25~ = 0.71, p = 0.41), or their interaction (F~3,75~ = 0.22, p = 0.88).
CD45+ counts are relatively stable across the cycle regardless of birth control use.

### Model 2: Neutrophils ~ time point × arm

```{r}
m2 <- lmer(log(neutrophils) ~ time_point * arm + (1 | pid), data = master)
anova(m2, type = 3) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  gt() |>
  tab_header(
    title = md("**Model 2.** log(Neutrophils) ~ time_point × arm"),
    subtitle = "Type III ANOVA (Satterthwaite)"
  ) |>
  fmt_number(columns = c("Sum Sq", "Mean Sq"), decimals = 3) |>
  fmt_number(columns = c("F value"), decimals = 2) |>
  fmt_number(columns = c("Pr(>F)"), decimals = 4)
```

**Result:** Same pattern — no significant effects of time point, arm, or their
interaction on neutrophil counts.

### Model 3: CD45+ ~ time point × PCOS status

```{r}
m3 <- lmer(log(cd45_positive) ~ time_point * pcos_status + (1 | pid), data = master)
anova(m3, type = 3) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  gt() |>
  tab_header(
    title = md("**Model 3.** log(CD45+) ~ time_point × PCOS status"),
    subtitle = "Type III ANOVA (Satterthwaite)"
  ) |>
  fmt_number(columns = c("Sum Sq", "Mean Sq"), decimals = 3) |>
  fmt_number(columns = c("F value"), decimals = 2) |>
  fmt_number(columns = c("Pr(>F)"), decimals = 4)
```

**Result:** No significant effects. PCOS status does not independently predict
CD45+ cell counts in this sample.

### Model 4: IL-1a ~ time point × arm **← the strongest finding**

```{r}
il1a_data <- luminex_master |>
  filter(cytokine == "IL-1a", limits == "not_censored")

m4 <- lmer(log(conc) ~ time_point * arm + (1 | pid), data = il1a_data)
anova(m4, type = 3) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  gt() |>
  tab_header(
    title = md("**Model 4.** log(IL-1a) ~ time_point × arm **"),
    subtitle = "Type III ANOVA (Satterthwaite)"
  ) |>
  fmt_number(columns = c("Sum Sq", "Mean Sq"), decimals = 3) |>
  fmt_number(columns = c("F value"), decimals = 2) |>
  fmt_number(columns = c("Pr(>F)"), decimals = 6)
```

**Result:** This is our strongest model. **IL-1a varies significantly across
the menstrual cycle** (F~3,73.8~ = 22.75, p < 0.001) and **differs by birth
control arm** (F~1,24.3~ = 6.96, p = 0.014). The interaction was not significant.
IL-1a peaks sharply at *onset* (menstruation) and at *end_bleeding*, with
lower concentrations in the birth control group.

### Model 5: CD4/CD8 ratio ~ time point × arm

```{r}
m5 <- lmer(log(cd4_cd8_ratio) ~ time_point * arm + (1 | pid), data = master)
anova(m5, type = 3) |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  gt() |>
  tab_header(
    title = md("**Model 5.** log(CD4/CD8 ratio) ~ time_point × arm"),
    subtitle = "Type III ANOVA (Satterthwaite)"
  ) |>
  fmt_number(columns = c("Sum Sq", "Mean Sq"), decimals = 3) |>
  fmt_number(columns = c("F value"), decimals = 2) |>
  fmt_number(columns = c("Pr(>F)"), decimals = 6)
```

**Result:** The CD4/CD8 ratio **changes significantly across cycle time
points** (F~3,75~ = 9.57, p < 0.001), but no effect of birth control arm or
their interaction. The ratio shifts across the cycle, suggesting differential
trafficking of helper vs. cytotoxic T cells during menstruation.

### Model summary table

```{r}
#| fig-cap: "Table 3. Model summary"
model_list <- list(
  "log(CD45+)" = m1,
  "log(Neutrophils)" = m2,
  "log(IL-1a)" = m4,
  "log(CD4/CD8)" = m5
)

model_rows <- list()
for (nm in names(model_list)) {
  aov <- anova(model_list[[nm]], type = 3)
  for (i in seq_len(nrow(aov))) {
    term <- rownames(aov)[i]
    model_rows[[length(model_rows) + 1]] <- tibble(
      Outcome = nm,
      Term = term,
      F = sprintf("%.2f", aov[i, "F value"]),
      df = paste0(aov[i, "NumDF"], ", ", round(aov[i, "DenDF"], 0)),
      p = sprintf("%.4f", aov[i, "Pr(>F)"]),
      Sig = case_when(
        aov[i, "Pr(>F)"] < 0.001 ~ "***",
        aov[i, "Pr(>F)"] < 0.01 ~ "**",
        aov[i, "Pr(>F)"] < 0.05 ~ "*",
        TRUE ~ "ns"
      )
    )
  }
}

bind_rows(model_rows) |>
  gt(groupname_col = "Outcome") |>
  tab_header(
    title = md("**Table 3.** Mixed-effects model results"),
    subtitle = "Type III ANOVA (Satterthwaite). Fixed effects; random intercept per participant."
  ) |>
  cols_label(Sig = "") |>
  tab_source_note(source_note = "*** p < 0.001, ** p < 0.01, * p < 0.05, ns = not significant")
```

---

## Figures (ggpubr-style comparisons)

### Figure 1 — CD45+ immune cells by birth control arm

```{r}
#| fig-cap: "CD45+ immune cells across the menstrual cycle, grouped by hormonal birth control use"
#| fig-height: 5
#| fig-width: 8

arm_colors <- c("birth_control" = "#E41A1C", "no_birth_control" = "#377EB8")

p1 <- master |>
  ggplot(aes(x = time_point, y = cd45_positive, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5, size = 1.2) +
  scale_fill_manual(values = arm_colors) +
  scale_color_manual(values = arm_colors) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "CD45+ immune cells across the menstrual cycle",
    subtitle = "By hormonal birth control use",
    x = "Menstrual cycle time point",
    y = "CD45+ cells",
    fill = "Study arm",
    color = "Study arm",
    caption = "p-values from t-test at each time point"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )

p1 + stat_compare_means(
  aes(group = arm),
  label = "p.format",
  method = "t.test",
  bracket.size = 0.3,
  size = 2.8,
  tip.length = 0.01
)
```

### Figure 2 — IL-1a by PCOS status

```{r}
#| fig-cap: "IL-1a concentration across the menstrual cycle, grouped by PCOS status"
#| fig-height: 5
#| fig-width: 8

pcos_colors <- c("no disease" = "#4DAF4A", "pcos" = "#984EA3")

il1a_plot_data <- luminex_master |>
  filter(cytokine == "IL-1a", limits == "not_censored")

p2 <- il1a_plot_data |>
  ggplot(aes(x = time_point, y = conc, fill = pcos_status)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = pcos_status), width = 0.15, alpha = 0.5, size = 1.2) +
  scale_fill_manual(values = pcos_colors) +
  scale_color_manual(values = pcos_colors) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "IL-1a concentration across the menstrual cycle",
    subtitle = "By PCOS status — only non-censored values",
    x = "Menstrual cycle time point",
    y = "IL-1a (pg/mL)",
    fill = "PCOS status",
    color = "PCOS status",
    caption = "p-values from t-test at each time point"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )

p2 + stat_compare_means(
  aes(group = pcos_status),
  label = "p.format",
  method = "t.test",
  bracket.size = 0.3,
  size = 2.8,
  tip.length = 0.01
)
```

### Figure 3 — CD4/CD8 ratio (both arm and PCOS)

```{r}
#| fig-cap: "CD4/CD8 T cell ratio across the menstrual cycle, faceted by PCOS status and coloured by birth control use"
#| fig-height: 5
#| fig-width: 10

master |>
  ggplot(aes(x = time_point, y = cd4_cd8_ratio, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.4, size = 1) +
  scale_fill_manual(values = arm_colors) +
  scale_color_manual(values = arm_colors) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  labs(
    title = "CD4/CD8 T cell ratio across the menstrual cycle",
    subtitle = "Faceted by PCOS status; colored by birth control use",
    x = "Menstrual cycle time point",
    y = "CD4+ / CD8+ ratio",
    fill = "Study arm",
    color = "Study arm"
  ) +
  facet_wrap(~ pcos_status) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top",
    strip.text = element_text(face = "bold")
  )
```

---

## Summary of key findings

| Finding | Evidence |
|---------|----------|
| **IL-1a varies dramatically across the cycle** | F~3,73.8~ = 22.75, p < 0.001 — peaks at onset and end_bleeding |
| **Birth control lowers IL-1a** | F~1,24.3~ = 6.96, p = 0.014 — lower concentrations in BC users |
| **CD4/CD8 ratio shifts across the cycle** | F~3,75~ = 9.57, p < 0.001 — suggests differential T cell trafficking |
| **CD45+ cells and neutrophils are stable** | No significant effects of time, arm, or PCOS |
| **PCOS shows trends but not significant** | Possibly due to small sample size (n = 27) |

## Limitations & next steps

- **Small sample (n = 27)** limits power for detecting interaction effects
- **Censored cytokine data** — TNFa, IFNg, IL-1b mostly below detection — reduces usable observations
- **Birth control type not specified** (pill vs. IUD vs. implant may differ)
- **PCOS diagnosis method** not documented
- Next: censored regression (tobit) models, inclusion of age/sexual activity
  as covariates, pathway analysis linking cytokines to cell types