---
title: "Session 1 — Exploring the Menstruation Dataset (Group B)"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  message: false
  warning: false
---

## Overview

This document loads and explores all four CSV files in the **menstruation
study** dataset (`data/group_b_menstruation/`). For each file we'll show:

-   What **one row** represents (the unit of observation)
-   What **every column** means
-   A quick glance at the data

```{r setup}
library(tidyverse)
```

---

## 00 — Sample IDs (`00_sample_ids_period.csv`)

```{r}
sample_ids <- read_csv("../data/group_b_menstruation/00_sample_ids_period.csv")
glimpse(sample_ids)
```

**What one row represents:**  
A single biological sample taken from one participant at one menstrual-cycle
time point.

**Column guide:**

| Column       | Type        | Meaning |
|--------------|-------------|---------|
| `pid`        | identifier  | Participant ID. Links to `01_participant_metadata_period.csv`. |
| `time_point` | categorical | When in the menstrual cycle the sample was taken — `week_prior` (week before period), `onset` (start of bleeding), `end_bleeding` (end of period), `week_post` (week after). |
| `arm`        | categorical | Study arm — `birth_control` (uses hormonal birth control) or `no_birth_control`. |
| `sample_id`  | identifier  | Unique sample code (e.g. SAMP001). Links to luminex and flow cytometry tables. |

```{r}
sample_ids |>
  count(arm, time_point) |>
  knitr::kable()
```

---

## 01 — Participant Metadata (`01_participant_metadata_period.csv`)

```{r}
metadata <- read_csv("../data/group_b_menstruation/01_participant_metadata_period.csv")
glimpse(metadata)
```

**What one row represents:**  
One study participant (27 participants total).

**Column guide:**

| Column          | Type        | Meaning |
|-----------------|-------------|---------|
| `pid`           | identifier  | Participant ID. Links to `00_sample_ids_period.csv`. |
| `arm`           | categorical | Study arm — `birth_control` or `no_birth_control`. |
| `age`           | numeric     | Age in years (range 26–41). |
| `pcos_status`   | categorical | Whether the participant has PCOS (polycystic ovary syndrome): `no disease` or `pcos`. |
| `period_product`| categorical | Menstrual product used: `tampon`, `pad`, or `menstrual_cup`. |
| `sex`           | categorical | All participants are `female`. |

```{r}
metadata |>
  count(arm, pcos_status) |>
  knitr::kable(col.names = c("Arm", "PCOS Status", "n"))

metadata |>
  summarise(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age)
  )
```

---

## 02 — Luminex Cytokine Results (`02_luminex_period.csv`)

```{r}
luminex <- read_csv("../data/group_b_menstruation/02_luminex_period.csv")
glimpse(luminex)

n_distinct(luminex$sample_id)  # how many unique samples
```

**What one row represents:**  
One cytokine measurement from one sample. The data is in "long" format —
every sample has 8 rows, one per cytokine. (108 samples × 8 cytokines = 864 rows.)

**Column guide:**

| Column      | Type        | Meaning |
|-------------|-------------|---------|
| `sample_id` | identifier  | Links to `00_sample_ids_period.csv`. |
| `cytokine`  | categorical | Immune signalling protein measured: IL-1a, IL-1b, IL-6, TNFa, MIG, IFNg, IP-10, MIP-3a. |
| `conc`      | numeric     | Concentration of the cytokine (pg/mL). |
| `limits`    | categorical | Reliability flag — `not_censored` (reliable), `below_detection_limit`, or `above_detect_limit` (outside the assay's measurable range). |

```{r}
# Which cytokines are measured
luminex |>
  count(cytokine) |>
  knitr::kable()

# Proportion of flagged results per cytokine
luminex |>
  mutate(flagged = limits != "not_censored") |>
  group_by(cytokine) |>
  summarise(pct_flagged = mean(flagged) * 100) |>
  arrange(desc(pct_flagged)) |>
  knitr::kable(digits = 1)
```

---

## 03 — Flow Cytometry Results (`03_flow_cytometry_period.csv`)

```{r}
flow <- read_csv("../data/group_b_menstruation/03_flow_cytometry_period.csv")
glimpse(flow)
```

**What one row represents:**  
Cell-count data from flow cytometry for one vaginal swab sample (108 samples
— one row per sample).

**Column guide:**

The column names trace a **gating hierarchy** — a process of narrowing down
cell types step by step:

```
All live cells
  ├── CD19-negative (non-B-cells)
  ├── CD45-negative (non-immune cells, e.g. epithelial)
  └── CD45-positive (immune cells / leukocytes)
        ├── Neutrophils (innate immune cells)
        └── Non-neutrophils (lymphocytes, macrophages, etc.)
              ├── CD3-negative (non-T-cells)
              └── CD3-positive (T cells)
                    ├── CD4+ helper T cells
                    └── CD8+ cytotoxic T cells
```

| Column              | Type    | Meaning |
|---------------------|---------|---------|
| `sample_id`         | identifier | Links to `00_sample_ids_period.csv`. |
| `live_cd19_negative` | numeric | Total viable cells that are *not* CD19+ (all live cells minus B cells). |
| `cd45_negative`     | numeric | Cells that are CD45- (non-immune cells, e.g. epithelial cells). |
| `cd45_positive`     | numeric | Cells that are CD45+ (immune cells / leukocytes). |
| `neutrophils`       | numeric | Neutrophils (a type of innate immune cell, counts within CD45+). |
| `non_neutrophils`   | numeric | CD45+ cells that are *not* neutrophils (lymphocytes, macrophages, etc.). |
| `cd3_negative`      | numeric | CD45+ cells that are CD3- (non-T-cells within the immune compartment). |
| `cd3_positive`      | numeric | CD45+ cells that are CD3+ (T cells). |
| `cd4_t_cells`       | numeric | CD4+ helper T cells (a subset of CD3+). |
| `cd8_t_cells`       | numeric | CD8+ cytotoxic T cells (a subset of CD3+). |

```{r}
# Median counts for key cell types
flow |>
  summarise(
    median_cd45_pos = median(cd45_positive),
    median_neutrophils = median(neutrophils),
    median_cd4 = median(cd4_t_cells),
    median_cd8 = median(cd8_t_cells)
  ) |>
  knitr::kable()
```

---

## How the tables link together

All four tables are linked through **key columns**:

```
01_participant_metadata  ─── pid ───→ 00_sample_ids
                                            │
                                     sample_id
                                        ├──→ 02_luminex_results  (8 rows per sample)
                                        └──→ 03_flow_cytometry   (1 row per sample)
```

-   `01_participant_metadata` ← `pid` → `00_sample_ids`  
    (one participant ↔ multiple samples across cycle time points)

-   `00_sample_ids` ← `sample_id` → `02_luminex_period`  
    (one sample ↔ 8 cytokine rows)

-   `00_sample_ids` ← `sample_id` → `03_flow_cytometry_period`  
    (one sample ↔ one flow-cytometry row)

```{r}
# Join all three data tables together for a complete picture
flow_with_metadata <- flow |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  select(sample_id, pid, time_point, arm, everything())

flow_with_metadata |>
  count(arm, time_point) |>
  knitr::kable()
```

### Arm vs. time point — immune cell patterns

```{r}
#| fig-height: 5
#| fig-width: 8

# Reorder time points so they plot in chronological order
flow_with_metadata <- flow_with_metadata |>
  mutate(time_point = fct_relevel(time_point,
    "week_prior", "onset", "end_bleeding", "week_post"))

# CD45+ immune cells across the cycle, coloured by study arm
flow_with_metadata |>
  ggplot(aes(x = time_point, y = cd45_positive, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "Immune cell (CD45+) counts across the menstrual cycle",
    subtitle = "By study arm — birth control vs. no birth control",
    x = "Time point",
    y = "CD45+ cells (log scale)",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

### Neutrophils — the most abundant immune cell

```{r}
#| fig-height: 5
#| fig-width: 8

flow_with_metadata |>
  ggplot(aes(x = time_point, y = neutrophils, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "Neutrophil counts across the menstrual cycle",
    subtitle = "By study arm — birth control vs. no birth control",
    x = "Time point",
    y = "Neutrophils (log scale)",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

### Key cytokine — IL-1a across the cycle

```{r}
#| fig-height: 5
#| fig-width: 8

# Join luminex data and pick a cytokine that's not mostly below detection
luminex_plot <- luminex |>
  filter(cytokine == "IL-1a", limits == "not_censored") |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  mutate(time_point = fct_relevel(time_point,
    "week_prior", "onset", "end_bleeding", "week_post"))

luminex_plot |>
  ggplot(aes(x = time_point, y = conc, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "IL-1a concentration across the menstrual cycle",
    subtitle = "By study arm — birth control vs. no birth control",
    x = "Time point",
    y = "IL-1a (pg/mL, log scale)",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

### T cells — CD4+/CD8+ ratio

```{r}
#| fig-height: 5
#| fig-width: 8

flow_with_metadata |>
  mutate(cd4_cd8_ratio = cd4_t_cells / cd8_t_cells) |>
  ggplot(aes(x = time_point, y = cd4_cd8_ratio, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    title = "CD4+/CD8+ T cell ratio across the menstrual cycle",
    subtitle = "Dashed line = equal numbers of helper and cytotoxic T cells",
    x = "Time point",
    y = "CD4 / CD8 ratio",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

---

## Correlation heatmap — immune cell counts × cytokines

A "bibliometric map" visualises how variables are associated with each other.
Here we compute pairwise correlations between all numeric variables (flow
cytometry cell counts × cytokine concentrations) and display them as a
coloured heatmap. Red = positive association, blue = negative, bigger dot =
stronger correlation.

```{r}
#| fig-height: 7
#| fig-width: 8

# Build a wide dataset: one row per sample, with flow + luminex side by side
# First pivot luminex so each cytokine is its own column
luminex_wide <- luminex |>
  filter(limits == "not_censored") |>
  select(sample_id, cytokine, conc) |>
  pivot_wider(names_from = cytokine, values_from = conc)

# Join with flow cytometry
cor_data <- flow |>
  left_join(luminex_wide, by = "sample_id") |>
  select(-sample_id) |>
  # Remove any rows with missing values
  drop_na()

# Compute the correlation matrix
cormat <- cor(cor_data, method = "spearman")

# Turn it into a tidy format for ggplot
cormat_tidy <- cormat |>
  as_tibble(rownames = "var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "rho")

cormat_tidy |>
  ggplot(aes(x = var1, y = var2, fill = rho, label = round(rho, 2))) +
  geom_tile(color = "white") +
  geom_text(size = 2.5, color = "grey30") +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-1, 1), name = "Spearman ρ"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Bibliometric association map",
    subtitle = "Spearman correlations between immune cell counts and cytokines",
    x = NULL, y = NULL
  )
```

---

## Quick notes

-   **Study design:** This is a longitudinal study following 27 participants
    across 4 time points in their menstrual cycle (`week_prior` → `onset` →
    `end_bleeding` → `week_post`). Half use birth control, half don't.

-   **PCOS:** About half the participants have PCOS, which is relevant
    because PCOS affects hormone levels and immune function.

-   **Missing / flagged data:** Luminex has `below_detection_limit` and
    `above_detect_limit` flags — these are censored values (too low or too
    high to read reliably). Some cytokines (like TNFa) are frequently below
    detection.

-   **Flow cytometry gating:** The column names trace a "gating hierarchy":
    live cells → CD45+ vs CD45- → neutrophils vs non-neutrophils → CD3+ vs
    CD3- → CD4+ vs CD8+ T cells.

-   **Things we don't know yet:** Which type of birth control (pill, IUD,
    implant)? How was PCOS diagnosed? What's the hypothesis about how birth
    control affects immune cells during the cycle?
