---
title: "Session 1 — Exploring the Yogurt Dataset (Group A)"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  message: false
  warning: false
---

## Overview

This document loads and explores all four CSV files in the **yogurt study**
dataset (`data/group_a_yogurt/`). For each file we'll show:

-   What **one row** represents (the unit of observation)
-   What **every column** means
-   A quick glance at the data

```{r setup}
library(tidyverse)
```

---

## 00 — Sample IDs (`00_sample_ids_yogurt.csv`)

```{r}
sample_ids <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids)
```

**What one row represents:**  
A single biological sample taken from one participant at one time point.

**Column guide:**

| Column       | Type        | Meaning |
|--------------|-------------|---------|
| `pid`        | identifier  | Participant ID. Links to `01_participant_metadata_yogurt.csv`. |
| `time_point` | categorical | When the sample was taken — `baseline` (before antibiotics) or `after_antibiotic` (after a course of antibiotics). |
| `arm`        | categorical | Study arm — `yogurt` (ate yogurt during antibiotics) or `unchanged_diet` (no yogurt, normal diet). |
| `sample_id`  | identifier  | Unique sample code (e.g. `UNCH001` or `YOG002`). Links to qPCR and luminex tables. |

```{r}
sample_ids |>
  count(arm, time_point) |>
  knitr::kable(col.names = c("Arm", "Time Point", "n"))
```

---

## 01 — Participant Metadata (`01_participant_metadata_yogurt.csv`)

```{r}
metadata <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(metadata)
```

**What one row represents:**  
One study participant (51 participants total).

**Column guide:**

| Column              | Type        | Meaning |
|---------------------|-------------|---------|
| `pid`               | identifier  | Participant ID. Links to `00_sample_ids_yogurt.csv`. |
| `arm`               | categorical | Study arm — `yogurt` or `unchanged_diet`. |
| `days_since_last_sex` | numeric   | Days since the participant last had sex (range 1–48). |
| `birth_control`     | categorical | Birth control method — `Depoprovera` (injectable contraceptive) or `no hormonal birth control`. |
| `age`               | numeric     | Age in years (range 24–37). |
| `education`         | categorical | Highest education level — `less than grade 9`, `grade 10-12, not matriculated`, `grade 10-12, matriculated`, or `post-secondary`. |
| `sex`               | categorical | All participants are `female`. |

```{r}
metadata |>
  count(arm, birth_control) |>
  knitr::kable(col.names = c("Arm", "Birth Control", "n"))

metadata |>
  summarise(
    n = n(),
    mean_age = mean(age),
    min_age = min(age),
    max_age = max(age)
  )
```

---

## 02 — qPCR Results (`02_qpcr_results_yogurt.csv`)

```{r}
qpcr <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr)
```

**What one row represents:**  
Quantitative PCR (qPCR) measurements of bacterial DNA from one vaginal swab
sample (102 samples total — one row per sample).

**Column guide:**

| Column          | Type    | Meaning |
|-----------------|---------|---------|
| `sample_id`     | identifier | Links to `00_sample_ids_yogurt.csv`. |
| `qpcr_bacteria` | numeric | Total bacterial load (16S rRNA gene copies). |
| `qpcr_crispatus`| numeric | Amount of *Lactobacillus crispatus* DNA — a "good" bacterium that helps protect against infections. |
| `qpcr_iners`    | numeric | Amount of *Lactobacillus iners* DNA — another *Lactobacillus* species, but less protective than *L. crispatus*. |

```{r}
# Quick look at the range of bacterial loads
qpcr |>
  summarise(
    min_bacteria = min(qpcr_bacteria),
    median_bacteria = median(qpcr_bacteria),
    max_bacteria = max(qpcr_bacteria),
    pct_detectable_crispatus = mean(qpcr_crispatus > 0) * 100,
    pct_detectable_iners = mean(qpcr_iners > 0) * 100
  ) |>
  knitr::kable(digits = 1)
```

---

## 03 — Luminex Cytokine Results (`03_luminex_results_yogurt.csv`)

```{r}
luminex <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex)

n_distinct(luminex$sample_id)  # how many unique samples
```

**What one row represents:**  
One cytokine measurement from one sample. The data is in "long" format —
every sample has 10 rows, one per cytokine. (102 samples × 10 cytokines = 1,020 rows.)

**Column guide:**

| Column      | Type        | Meaning |
|-------------|-------------|---------|
| `sample_id` | identifier  | Links to `00_sample_ids_yogurt.csv`. |
| `cytokine`  | categorical | Immune signalling protein measured: IL-1a, IL-1b, IL-6, IL-8, IL-10, TNFa, IFNg (listed as `IFN-Y`), IP-10, MIG, MIP-3a. |
| `conc`      | numeric     | Concentration of the cytokine (pg/mL). |
| `limits`    | categorical | Reliability flag — `within limits` (reliable reading) or `out of range` (concentration too low or too high for the assay to measure accurately). |

```{r}
# Which cytokines are measured
luminex |>
  count(cytokine) |>
  knitr::kable()

# Proportion of out-of-range results per cytokine
luminex |>
  mutate(flagged = limits != "within limits") |>
  group_by(cytokine) |>
  summarise(pct_flagged = mean(flagged) * 100) |>
  arrange(desc(pct_flagged)) |>
  knitr::kable(digits = 1)
```

---

## How the tables link together

All four tables are linked through **key columns**:

```
01_participant_metadata  ─── pid ───→ 00_sample_ids
                                            │
                                     sample_id
                                        ├──→ 02_qpcr_results    (1 row per sample)
                                        └──→ 03_luminex_results (10 rows per sample)
```

-   `01_participant_metadata` ← `pid` → `00_sample_ids`  
    (one participant ↔ multiple samples: baseline + after antibiotic)

-   `00_sample_ids` ← `sample_id` → `02_qpcr_results`  
    (one sample ↔ one qPCR row)

-   `00_sample_ids` ← `sample_id` → `03_luminex_results`  
    (one sample ↔ 10 cytokine rows)

```{r}
# Join all three data tables together for a complete picture
qpcr_with_metadata <- qpcr |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  select(sample_id, pid, time_point, arm, everything())

qpcr_with_metadata |>
  count(arm, time_point) |>
  knitr::kable()
```

### Arm vs. time point — total bacterial load

```{r}
#| fig-height: 5
#| fig-width: 8

qpcr_with_metadata <- qpcr_with_metadata |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

# Total bacteria across time points, by study arm
qpcr_with_metadata |>
  ggplot(aes(x = time_point, y = qpcr_bacteria, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "Total bacterial load across time points",
    subtitle = "By study arm — yogurt vs. unchanged diet",
    x = "Time point",
    y = "Total bacteria (log scale)",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

### L. crispatus vs. L. iners — which species dominates?

```{r}
#| fig-height: 5
#| fig-width: 8

# Add a small constant so we can plot zeros on a log scale
qpcr_with_metadata |>
  ggplot(aes(x = qpcr_crispatus + 1, y = qpcr_iners + 1, color = arm)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  geom_abline(slope = 1, linetype = "dashed", color = "grey50") +
  labs(
    title = "L. crispatus vs. L. iners across all samples",
    subtitle = "Dashed line = equal amounts. Points above it = more iners, below = more crispatus",
    x = "L. crispatus + 1 (log scale)",
    y = "L. iners + 1 (log scale)",
    color = "Arm"
  ) +
  theme_minimal()
```

### Key cytokine — IL-1a across time points

```{r}
#| fig-height: 5
#| fig-width: 8

# Join luminex data and pick IL-1a
luminex_plot <- luminex |>
  filter(cytokine == "IL-1a") |>
  left_join(sample_ids, by = "sample_id") |>
  left_join(metadata, by = "pid", suffix = c(".sample", "")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

luminex_plot |>
  ggplot(aes(x = time_point, y = conc, fill = arm)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(aes(color = arm), width = 0.15, alpha = 0.5) +
  scale_y_log10() +
  labs(
    title = "IL-1a concentration across time points",
    subtitle = "By study arm — yogurt vs. unchanged diet",
    x = "Time point",
    y = "IL-1a (pg/mL, log scale)",
    fill = "Arm",
    color = "Arm"
  ) +
  theme_minimal()
```

---

## Quick notes

-   **Study design:** This is a longitudinal study following 51 participants
    at 2 time points — `baseline` (before antibiotics) and
    `after_antibiotic`. Half ate yogurt during their antibiotic course, the
    other half kept their normal diet.

-   **Why yogurt?** The study is testing whether eating yogurt (which
    contains live bacteria) helps restore a healthy vaginal microbiome after
    antibiotics disrupt it.

-   **qPCR targets:** The study measures *Lactobacillus crispatus* (a "good"
    species that produces protective lactic acid) and *Lactobacillus iners*
    (a related species that's less protective). Many samples have zero
    *L. crispatus* — it was undetectable.

-   **Birth control:** About half the participants use Depoprovera (an
    injectable hormonal contraceptive), which can affect the vaginal
    microbiome.

-   **Missing / flagged data:** Luminex has `out of range` flags. Some
    cytokines (like IL-10, TNFa) are frequently out of range, meaning they're
    too low to measure reliably in many samples.

-   **Things we don't know yet:** What specific antibiotic was given? For how
    long? How much yogurt did the yogurt group eat? Were there any side
    effects?
