---
title: "Module 6 — Demographics and Table 1"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi and let it write the code. Group B: **manual** — write each step yourself. A chat LLM may answer conceptual questions but should not edit this notebook or run commands. Exercise 2 flips the modes.

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)

table_01 <- read_csv(
  here("data/instructional_dataset/01_participant_metadata_UKZN_workshop_2023.csv"),
  col_types = cols(sex = col_character())
)
```

## Part 1 — Predict (no code)

```{r}
table_01
```

For each non-pid column of `table_01` (`arm`, `smoker`, `age`, `education`, `sex`), predict in plain words below. Don't run any code yet.

**1. `arm`**

- How will `tbl_summary(by = arm)` summarise it? *Your answer:*
it will catagorize based on the two groups (control and intervention) and give counts and percentages for each group.
- What test will `add_p()` pick? *Your answer:*
maybe a chi-squared test, since it's a categorical variable with two groups.
**2. `smoker`** 
catagorical variable with two groups (smoker and non-smoker), so it will give counts and percentages for each group.

- *Your answer (summary, test):*
chi-squared test, since it's a categorical variable with two groups.
**3. `age`**
continuous variable, so it will give mean and standard deviation for each group.
- *Your answer (summary, test):*
mean (SD), since it's a continuous variable.
**4. `education`**
catagorical variable with multiple groups (primary, secondary, tertiary), so it will give counts and percentages for each group.
- *Your answer (summary, test):*
chi-squared test, since it's a categorical variable with multiple groups.

**5. `sex`**
totalnumber
- *Your answer (summary, test):*
it will give counts and percentages for each group, but since it's not by a group, it will just give the overall distribution

## Part 2 — Build

Each step adds one piece on top of the previous table. Run after each step.

```{r}
# step 1: tbl_summary defaults
table_01 |>
  select(-pid, -sex) |>
  tbl_summary(by = arm)
```

```{r}
# step 2: add p-values
# (modify the previous chunk by adding `|> add_p()`)
table_01 |> select(-pid, -sex) |> tbl_summary(by = arm) |> add_p()
```

```{r}
# step 3: override age to mean (SD) + t-test
statistic = list(age ~ "{mean} ({sd})")
test = list(age ~ "t.test")
```

```{r}
# step 4: polish
# add `|> add_overall() |> bold_labels() |> modify_header(label ~ "**Characteristic**")`
```

## Part 3 — Reach (optional)

```{r}
# reach: same table, but by = smoker instead of by = arm
# what question is this table now answering?
```
